---
title: "Importing and Exporting Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Importing and Exporting Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Survey data lives in many formats --- SPSS (.sav), Stata (.dta), SAS (.sas7bdat), and Excel (.xlsx). mariposa reads all of them and preserves the metadata that makes survey data special: variable labels, value labels, and missing value definitions.

| Function | Format | Notes |
|----------|--------|-------|
| `read_spss()` | SPSS .sav | Tagged NA support for user-defined missing values |
| `read_por()` | SPSS .por | Portable SPSS format |
| `read_stata()` | Stata .dta | Extended missing values |
| `read_sas()` | SAS .sas7bdat | Catalog file support for labels |
| `read_xpt()` | SAS .xpt | SAS transport format |
| `read_xlsx()` | Excel .xlsx | Label reconstruction from metadata sheets |

For export, mariposa writes data back to statistical formats with full preservation of labels and missing values:

| Function | Format | Notes |
|----------|--------|-------|
| `write_spss()` | SPSS .sav | Tagged NA roundtripping |
| `write_stata()` | Stata .dta | Label preservation |
| `write_xpt()` | SAS .xpt | SAS transport format |
| `write_xlsx()` | Excel .xlsx | Multiple export modes (data, codebook, frequencies) |

## Reading SPSS Files

### Basic Import

SPSS is the most common format in survey research. `read_spss()` reads .sav files and preserves all metadata:

```{r, eval = FALSE}
# Read an SPSS file
data <- read_spss("survey_2024.sav")
```

The result is a tibble with `haven_labelled` columns. Each column carries its variable label, value labels, and missing value definitions as attributes.

### Tagged Missing Values

SPSS allows researchers to define multiple types of missing values (e.g., -9 = "refused", -8 = "don't know", -7 = "not applicable"). By default, `read_spss()` preserves these as *tagged NAs* --- special NA values that remember which missing code they came from:

```{r, eval = FALSE}
# Tagged NAs are on by default
data <- read_spss("survey.sav", tag.na = TRUE)

# Turn off if you just want regular NAs
data <- read_spss("survey.sav", tag.na = FALSE)
```

### Working with Tagged NAs

Once imported, you can inspect the tagged NAs:

```{r, eval = FALSE}
# See a breakdown of missing types
na_frequencies(data, q1, q2, q3)

# Convert tagged NAs back to their original codes
data_with_codes <- untag_na(data, q1, q2)

# Remove tags but keep NAs
data_clean <- strip_tags(data)
```

`na_frequencies()` is particularly useful for understanding response patterns --- it shows you how many respondents refused, said "don't know", or skipped each question.

## Reading Other Formats

### Stata

```{r, eval = FALSE}
data <- read_stata("survey.dta")
```

Stata files carry variable labels and value labels, similar to SPSS. Extended missing values (.a through .z) are preserved.

### SAS

```{r, eval = FALSE}
# SAS data file
data <- read_sas("survey.sas7bdat")

# With a separate catalog file for labels
data <- read_sas("survey.sas7bdat", catalog_file = "formats.sas7bcat")

# SAS transport format
data <- read_xpt("survey.xpt")
```

### SPSS Portable

```{r, eval = FALSE}
data <- read_por("survey.por")
```

### Excel

```{r, eval = FALSE}
# Basic Excel import
data <- read_xlsx("survey.xlsx")

# Excel file with label metadata (exported by write_xlsx)
data <- read_xlsx("survey.xlsx", label_sheet = "labels")
```

`read_xlsx()` can reconstruct variable and value labels from a metadata sheet, enabling a full roundtrip through Excel format.

## Inspecting Imported Data

After importing, use `codebook()` to get an interactive overview of your data:

```{r, eval = FALSE}
codebook(survey_data)
```

The codebook displays in the RStudio Viewer and shows:

- Variable names, types, and positions
- Variable labels (descriptions)
- Value labels with frequency counts
- Missing value breakdown (tagged NAs)
- Empirical value ranges for numeric variables

For a quick look at variable names and labels without leaving the console, use `find_var()`:

```{r}
# Find variables related to "trust"
find_var(survey_data, "trust")

# Search by variable label
find_var(survey_data, "satisfaction", search = "label")
```

## Exporting Data

### Writing SPSS Files

`write_spss()` creates .sav files with full preservation of labels and missing values. If your data contains tagged NAs (from `read_spss()`), they are converted back to SPSS user-defined missing values:

```{r, eval = FALSE}
# Basic export
write_spss(survey_data, "output.sav")

# With compression options
write_spss(survey_data, "output.sav", compress = "zsav")  # smaller file
```

### Writing Stata Files

```{r, eval = FALSE}
write_stata(survey_data, "output.dta")
```

### Writing SAS Transport Files

```{r, eval = FALSE}
write_xpt(survey_data, "output.xpt")
```

### Writing Excel Files

`write_xlsx()` supports multiple export modes:

```{r, eval = FALSE}
# Export data only
write_xlsx(survey_data, "output.xlsx")

# Export a codebook
cb <- codebook(survey_data)
write_xlsx(cb, "codebook.xlsx")

# Export frequency tables
freq <- frequency(survey_data, education, employment)
write_xlsx(freq, "frequencies.xlsx")
```

## Roundtripping: SPSS → R → SPSS

A key strength of mariposa is lossless roundtripping. Data imported from SPSS can be exported back without losing any information:

```{r, eval = FALSE}
# 1. Import from SPSS
original <- read_spss("survey.sav")

# 2. Work with the data in R
processed <- original %>%
  filter(age >= 18) %>%
  mutate(age_group = rec(., age, rules = "18:29=1; 30:49=2; 50:99=3"))

# 3. Export back to SPSS
write_spss(processed, "survey_processed.sav")

# Variable labels, value labels, and missing value definitions
# are all preserved in the exported file.
```

## Practical Tips

1. **Always use `read_spss()` over `haven::read_sav()`** when your SPSS file uses user-defined missing values. The tagged NA system preserves the distinction between "refused" and "don't know" responses.

2. **Inspect before analyzing.** After import, run `codebook()` or `find_var()` to understand what variables are available and how they are coded.

3. **Keep tagged NAs during analysis.** All mariposa functions handle tagged NAs correctly --- they are treated as missing values in computations but retain their type information.

4. **Use `strip_tags()` when handing data to non-mariposa functions.** Some R functions may not handle tagged NAs correctly. Strip the tags first to convert them to regular NAs.

5. **Prefer `write_spss()` for SPSS users.** The roundtripping ensures your colleagues can open the file in SPSS with all metadata intact.

## Summary

1. `read_spss()`, `read_stata()`, `read_sas()`, `read_xlsx()` import data with full metadata preservation
2. Tagged NAs preserve SPSS user-defined missing value types
3. `codebook()` and `find_var()` help you understand imported data quickly
4. `write_spss()`, `write_stata()`, `write_xpt()`, `write_xlsx()` export with label and missing value roundtripping
5. The full import-export cycle is lossless for SPSS data

## Next Steps

- Learn how to work with labels and missing values --- see `vignette("labels-and-missing-values")`
- Transform and recode variables --- see `vignette("data-transformation")`
- Start analyzing your data --- see `vignette("descriptive-statistics")`
