---
title: "Descriptive Statistics and Frequencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Descriptive Statistics and Frequencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Descriptive statistics help you understand your data before doing any complex analysis. Think of it as getting to know your survey responses --- what is typical, what is unusual, and how spread out the answers are.

This guide covers three functions:

- `describe()` --- Summary statistics for numeric variables
- `frequency()` --- Frequency tables for categorical variables
- `crosstab()` --- Cross-tabulations comparing two categorical variables

## Getting Summary Statistics

### Basic Usage

The simplest way to get descriptive statistics:

```{r}
survey_data %>% describe(age)
```

This gives you the mean ($\bar{x}$), median, standard deviation (*SD*), and other key statistics.

### Multiple Variables at Once

Analyze several variables together:

```{r}
survey_data %>%
  describe(age, income, life_satisfaction)
```

### With Survey Weights

To get population-representative statistics, add the `weights` argument:

```{r}
survey_data %>%
  describe(age, income, life_satisfaction,
           weights = sampling_weight)
```

The weighted results better represent your target population by correcting for sampling biases.

### Grouped Analysis

Compare statistics across groups using `group_by()`:

```{r}
survey_data %>%
  group_by(region) %>%
  describe(age, income, weights = sampling_weight)
```

### Customizing Output

Choose which statistics to display with the `show` argument:

```{r}
# Just the essentials
survey_data %>%
  describe(age, income, show = c("mean", "sd", "range"))

# Everything available
survey_data %>%
  describe(age, show = "all")
```

## Frequency Tables

### Basic Frequency

For categorical variables, use `frequency()`:

```{r}
survey_data %>%
  frequency(education)
```

### Understanding the Output

The output includes:

- **Frequency**: Count of responses in each category
- **Percent**: Percentage of all cases (including missing)
- **Valid Percent**: Percentage of non-missing cases only
- **Cumulative Percent**: Running total of valid percentages

When reporting results, use *Valid Percent* if missing values are truly missing (e.g., skip patterns), and *Percent* when non-response is meaningful.

### Weighted Frequencies

```{r}
survey_data %>%
  frequency(education, weights = sampling_weight)
```

### Multiple Variables

Analyze several categorical variables at once:

```{r}
survey_data %>%
  frequency(education, employment, region,
            weights = sampling_weight)
```

### Grouped Frequencies

See how distributions vary by group:

```{r}
survey_data %>%
  group_by(gender) %>%
  frequency(education, weights = sampling_weight)
```

## Cross-Tabulation

### Basic Crosstab

Examine the relationship between two categorical variables:

```{r}
survey_data %>%
  crosstab(education, employment)
```

### Understanding the Output

Cross-tabulations show several types of percentages:

- **Count**: Number of cases in each cell
- **Row %**: Percentage within each row (sums to 100% across)
- **Column %**: Percentage within each column (sums to 100% down)
- **Cell %**: Percentage of the total sample

Row percentages answer: "Of people with this education level, what proportion has each employment status?"
Column percentages answer: "Of people with this employment status, what proportion has each education level?"

### Weighted Crosstabs

```{r}
survey_data %>%
  crosstab(education, employment, weights = sampling_weight)
```

### Grouped Crosstabs

Create separate cross-tabulations for each subgroup:

```{r}
survey_data %>%
  group_by(region) %>%
  crosstab(education, employment, weights = sampling_weight)
```

## Practical Tips

### 1. Always Check Your Data First

Start with basic descriptives to spot potential issues:

```{r}
survey_data %>%
  describe(age, income, show = "all")
```

Look for:

- **Impossible values** (e.g., negative ages) indicating data entry errors
- **High rates of missing data** that might bias results
- **Extreme skewness** (values beyond $\pm 1$) suggesting non-normal distributions

### 2. Use Weights When Available

Unweighted results can be misleading:

```{r}
unweighted_mean <- survey_data %>%
  summarise(mean_age = mean(age, na.rm = TRUE)) %>%
  pull()

weighted_result <- w_mean(survey_data, age, weights = sampling_weight)

cat("Unweighted mean age:", round(unweighted_mean, 1), "\n")
cat("Weighted mean age:", round(weighted_result$results$weighted_mean, 1), "\n")
```

### 3. Consider Your Audience

- **Technical audience**: Include *SD*, skewness, kurtosis
- **General audience**: Focus on mean, median, and percentages

## Complete Example

A typical descriptive analysis workflow:

```{r}
# 1. Overall summary
survey_data %>%
  describe(age, income, life_satisfaction,
           weights = sampling_weight,
           show = "short")

# 2. Key distributions
survey_data %>%
  frequency(education, employment,
            weights = sampling_weight)

# 3. Relationship between categories
survey_data %>%
  crosstab(education, employment,
           weights = sampling_weight)

# 4. Regional comparisons
survey_data %>%
  group_by(region) %>%
  describe(income, life_satisfaction,
           weights = sampling_weight)
```

## Summary

1. **`describe()`** gives you a full picture of numeric variables
2. **`frequency()`** shows the distribution of categorical variables
3. **`crosstab()`** reveals relationships between two categorical variables
4. All three functions support **survey weights** and **grouped analysis**
5. Always explore your data descriptively before running statistical tests

## Next Steps

- Test for significant differences with `t_test()` or `chi_square()` --- see `vignette("hypothesis-testing")`
- Explore relationships with `pearson_cor()` --- see `vignette("correlation-analysis")`
- Learn about weighted analysis --- see `vignette("survey-weights")`
