---
title: "Comparing Groups and Testing Hypotheses"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Groups and Testing Hypotheses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Statistical tests help you determine whether differences between groups are real or just due to random chance. This guide covers the main tests available in mariposa:

| Test | Use when... |
|------|-------------|
| `t_test()` | Comparing means between **two** groups |
| `oneway_anova()` | Comparing means across **three or more** groups |
| `chi_square()` | Testing relationships between **categorical** variables |
| `mann_whitney()` | Non-parametric alternative when data is not normally distributed |

## t-Tests

### Independent Samples t-Test

Compare average life satisfaction between genders:

```{r}
survey_data %>%
  t_test(life_satisfaction, group = gender)
```

With survey weights for representative results:

```{r}
survey_data %>%
  t_test(life_satisfaction, group = gender, weights = sampling_weight)
```

The output shows two versions: one assuming equal variances (Student's t-test) and one not assuming equal variances (Welch's t-test). When in doubt, use the Welch version --- it is more robust.

### Multiple Variables at Once

Test several outcome variables simultaneously:

```{r}
survey_data %>%
  t_test(trust_government, trust_science, trust_media,
         group = gender, weights = sampling_weight)
```

### One-Sample t-Test

Test whether a population mean differs from a specific value. For example, is average life satisfaction different from 3 (the scale midpoint)?

```{r}
survey_data %>%
  t_test(life_satisfaction, mu = 3, weights = sampling_weight)
```

### Grouped Analysis

Run separate tests for each region:

```{r}
survey_data %>%
  group_by(region) %>%
  t_test(income, group = gender, weights = sampling_weight)
```

## One-Way ANOVA

### Basic ANOVA

Compare means across multiple groups (3+ categories):

```{r}
survey_data %>%
  oneway_anova(life_satisfaction, group = education)
```

### With Survey Weights

```{r}
survey_data %>%
  oneway_anova(life_satisfaction, group = education,
               weights = sampling_weight)
```

The output includes $\eta^2$ (eta-squared), which tells you how much of the variance in the outcome is explained by group membership:

- **Small effect**: $\eta^2 \approx 0.01$
- **Medium effect**: $\eta^2 \approx 0.06$
- **Large effect**: $\eta^2 \approx 0.14$

### Post-Hoc Tests

When ANOVA is significant, it tells you that *at least one* group differs --- but not *which* groups differ. Use post-hoc tests to find out:

```{r}
# Save the ANOVA result
anova_result <- survey_data %>%
  oneway_anova(life_satisfaction, group = education,
               weights = sampling_weight)

# Tukey HSD: identifies which pairs of groups differ
tukey_test(anova_result)

# Scheffe test: more conservative (fewer false positives)
scheffe_test(anova_result)
```

### Testing Assumptions

ANOVA assumes equal variances across groups. Check this with Levene's test:

```{r}
levene_test(anova_result)
```

If Levene's test is significant ($p < 0.05$), variances are unequal. In that case, the Welch correction (included in the ANOVA output) is more appropriate.

## Chi-Square Test

### Basic Chi-Square

Test whether two categorical variables are related:

```{r}
survey_data %>%
  chi_square(education, employment)
```

A significant result means the variables are not independent --- knowing someone's education tells you something about their employment status (or vice versa).

### With Survey Weights

```{r}
survey_data %>%
  chi_square(education, employment, weights = sampling_weight)
```

### Multiple Comparisons

Test several relationships:

```{r}
survey_data %>%
  chi_square(employment, education, weights = sampling_weight)

survey_data %>%
  chi_square(employment, gender, weights = sampling_weight)
```

## Mann-Whitney Test

### When to Use

Use the Mann-Whitney *U* test (also called Wilcoxon rank-sum test) when:

- Your data is **not normally distributed**
- Sample sizes are **small**
- The variable is **ordinal** (ranked categories)

It compares the *distributions* rather than the means.

### Basic Usage

```{r}
survey_data %>%
  mann_whitney(political_orientation, group = region)
```

### With Survey Weights

```{r}
survey_data %>%
  mann_whitney(political_orientation, group = region,
               weights = sampling_weight)
```

## Interpreting Results

### Understanding p-values

- $p < 0.05$: The difference is statistically significant at the 5% level
- $p \geq 0.05$: No significant difference detected

Important: "not significant" does **not** mean "no difference" --- it means we cannot rule out that the difference is due to chance, given the sample size.

### Effect Sizes Matter

With large samples, even tiny differences can be statistically significant. Always check effect sizes:

```{r}
result <- survey_data %>%
  oneway_anova(income, group = education, weights = sampling_weight)
print(result)
```

The t-test provides Cohen's *d* (small: 0.2, medium: 0.5, large: 0.8) and ANOVA provides $\eta^2$.

### Sample Size Effects

Large samples detect small effects. Small samples may miss real effects. Consider:

- Always report effect sizes alongside p-values
- Look at confidence intervals for practical significance
- Do not interpret $p = 0.049$ and $p = 0.051$ as fundamentally different

## Complete Example

A typical hypothesis testing workflow:

```{r}
# 1. Descriptive overview
survey_data %>%
  group_by(education) %>%
  describe(life_satisfaction, weights = sampling_weight)

# 2. Test for overall differences
anova_result <- survey_data %>%
  oneway_anova(life_satisfaction, group = education,
               weights = sampling_weight)
print(anova_result)

# 3. Post-hoc: which groups differ?
tukey_test(anova_result)

# 4. Check assumptions
levene_test(anova_result)
```

## Best Practices

1. **Check assumptions first.** Inspect skewness and kurtosis before choosing a test. For non-normal data, consider `mann_whitney()`.
2. **Use appropriate tests.** Normal data with equal variances: t-test/ANOVA. Non-normal or ordinal: Mann-Whitney. Categorical: chi-square.
3. **Report completely.** Always include the test statistic, degrees of freedom, p-value, effect size, and confidence intervals.
4. **Watch for multiple comparisons.** Running many tests inflates the chance of false positives. Consider Bonferroni correction or pre-register your hypotheses.
5. **Significance is not importance.** A statistically significant result with a negligible effect size may not be practically relevant.

## Summary

1. **`t_test()`** compares means between two groups --- use it for questions like "Do men and women differ?"
2. **`oneway_anova()`** extends this to three or more groups, with post-hoc tests to identify which groups differ
3. **`chi_square()`** tests whether categorical variables are related
4. **`mann_whitney()`** is the non-parametric alternative when assumptions are not met
5. Always report **effect sizes** alongside p-values

## Next Steps

- Explore relationships between continuous variables --- see `vignette("correlation-analysis")`
- Learn about weighted analysis --- see `vignette("survey-weights")`
- Revisit your data --- see `vignette("descriptive-statistics")`
