---
title: "Introduction to SurveyStat"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to SurveyStat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SurveyStat)
library(dplyr)
```

## Overview

SurveyStat is designed to provide comprehensive tools for statistical analysis of survey data. This vignette introduces the basic functionality of the package.

## Basic Usage

### Descriptive Statistics

The package provides comprehensive descriptive statistics for survey data:

```{r}
# Load the synthetic survey data
data(survey_data)

# Basic descriptive statistics
survey_data %>% describe(age, income, weights = sampling_weight)

# Weighted mean calculation
survey_data %>% w_mean(life_satisfaction, weights = sampling_weight)
```

### Frequency Analysis

Analyze categorical variables with proper weighting:

```{r}
# Frequency analysis with weights
survey_data %>% frequency(education, region, weights = sampling_weight)
```

### Statistical Tests

Perform hypothesis tests with survey data:

```{r}
# Group comparison with t-test (male vs female)
survey_data %>% 
  t_test(life_satisfaction, group = gender, weights = sampling_weight)

# Or grouped analysis by region
survey_data %>%
  group_by(region) %>%
  t_test(life_satisfaction, group = gender, weights = sampling_weight)
```

## Package Features

SurveyStat provides a comprehensive suite of statistical tools:

- **Descriptive Statistics**: `describe()`, `frequency()`, weighted w_* functions
- **Hypothesis Testing**: t-tests, ANOVA, non-parametric tests  
- **Post-Hoc Analysis**: Tukey tests, estimated marginal means
- **Survey-Weighted Analysis**: Proper handling of sampling weights
- **Synthetic Datasets**: License-free data for testing and examples

## Getting Help

For questions and bug reports, please visit our [GitHub repository](https://github.com/yourusername/SurveyStat).