---
title: "Labels and Missing Values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Labels and Missing Values}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Survey data from SPSS, Stata, and SAS stores two kinds of labels:

- **Variable labels** describe what a variable measures (e.g., `gender` → "Respondent's gender")
- **Value labels** map numeric codes to text (e.g., `1` → "Male", `2` → "Female")

In R, these labels are stored as attributes on `haven_labelled` columns. mariposa provides 10 functions for inspecting, modifying, and converting labelled data --- plus tools for declaring missing values and searching variables.

| Function | Purpose |
|----------|---------|
| `var_label()` | Get or set variable labels |
| `val_labels()` | Get or set value labels |
| `find_var()` | Search variables by name or label pattern |
| `to_label()` | Convert labelled → factor |
| `to_character()` | Convert labelled → character |
| `to_numeric()` | Convert factor/labelled → numeric |
| `to_labelled()` | Convert factor/character → labelled |
| `set_na()` | Declare values as missing (tagged NAs) |
| `unlabel()` | Strip all label metadata |
| `copy_labels()` | Restore labels after dplyr operations |
| `drop_labels()` | Remove unused value labels |

## Inspecting Labels

### Variable Labels

Variable labels describe what each column contains. Use `var_label()` to retrieve them:

```{r}
# Get labels for specific variables
var_label(survey_data, gender, education, life_satisfaction)
```

```{r}
# Get labels for all variables
var_label(survey_data)
```

### Value Labels

Value labels map numeric codes to meaningful text. Use `val_labels()` to retrieve them:

```{r}
# Get value labels for a single variable
val_labels(survey_data, gender)
```

```{r}
# Get value labels for multiple variables
val_labels(survey_data, education, employment)
```

### Finding Variables

SPSS datasets often have cryptic variable names like `q104a_1` or `v23`. Use `find_var()` to search by name or label:

```{r}
# Search in both names and labels (default)
find_var(survey_data, "trust")
```

```{r}
# Search only in variable labels
find_var(survey_data, "satisfaction", search = "label")
```

```{r}
# Search only in variable names
find_var(survey_data, "age|income", search = "name")
```

## Setting Labels

### Setting Variable Labels

```{r}
# Set labels for specific variables
labeled_data <- var_label(survey_data,
  age = "Age of respondent in years",
  income = "Monthly net income in euros"
)

# Verify
var_label(labeled_data, age, income)
```

### Setting Value Labels

```{r}
# Set value labels
labeled_data <- val_labels(survey_data,
  gender = c("Male" = 1, "Female" = 2)
)

# Verify
val_labels(labeled_data, gender)
```

```{r}
# Add labels without replacing existing ones
labeled_data <- val_labels(survey_data,
  gender = c("Diverse" = 3),
  .add = TRUE
)

val_labels(labeled_data, gender)
```

## Converting Between Formats

Survey data often needs conversion between labelled, factor, character, and numeric formats depending on the analysis.

### Labelled → Factor

Use `to_label()` when you need factors for plotting or statistical models:

```{r}
# Convert gender from labelled to factor
factor_data <- to_label(survey_data, gender, education)

# Check the result
class(factor_data$gender)
levels(factor_data$gender)
head(factor_data$gender)
```

```{r}
# Create ordered factors (useful for ordinal variables)
ordered_data <- to_label(survey_data, education, ordered = TRUE)
levels(ordered_data$education)
```

### Labelled → Character

Use `to_character()` for string-based operations:

```{r}
char_data <- to_character(survey_data, gender, region)
head(char_data$gender)
```

### Factor → Numeric

Use `to_numeric()` to convert factors or labelled vectors back to numbers:

```{r}
# First convert to factor, then back to numeric
factor_data <- to_label(survey_data, education)

# Parse numeric values from factor levels
numeric_data <- to_numeric(factor_data, education)
head(numeric_data$education)
```

### Numeric/Factor → Labelled

Use `to_labelled()` to add labels to plain numeric or factor columns:

```{r}
# Convert a factor back to haven_labelled
plain_data <- data.frame(
  gender = factor(c("Male", "Female", "Male")),
  score = c(3, 4, 2)
)

labelled_data <- to_labelled(plain_data, gender)
class(labelled_data$gender)
```

## Declaring Missing Values

### Setting Values as Missing

Use `set_na()` to declare specific numeric codes as missing values:

```{r, eval = FALSE}
# After importing SPSS data where -9 = refused, -8 = don't know
data <- read_spss("survey.sav", tag.na = FALSE)

# Declare -9 and -8 as missing across all numeric columns
data <- set_na(data, -9, -8, tag = TRUE)

# Declare different missing codes for specific variables
data <- set_na(data, q1 = c(-9, -8), q2 = c(99))
```

When `tag = TRUE` (the default), each missing code becomes a distinct tagged NA, so you can later distinguish "refused" from "don't know" responses.

### Removing All Labels

Use `unlabel()` when you need plain numeric data without any label metadata:

```{r, eval = FALSE}
# Strip all labels from entire dataset
plain_data <- unlabel(survey_data)

# Strip labels from specific variables only
plain_data <- unlabel(survey_data, gender, education)
```

This converts `haven_labelled` columns to plain `numeric` or `character`, tagged NAs to regular `NA`, and removes all label attributes.

## Preserving Labels Through Pipelines

### The Problem

dplyr operations like `filter()`, `mutate()`, and `select()` can strip label attributes from columns:

```{r}
filtered <- survey_data %>%
  filter(age >= 30) %>%
  select(gender, education, income)

# Labels may be lost after certain operations
```

### The Solution: copy_labels()

Use `copy_labels()` to restore labels from the original data:

```{r}
# Restore labels from the source dataset
filtered <- copy_labels(filtered, survey_data)

# Verify labels are back
var_label(filtered, gender, education)
```

### Cleaning Unused Labels

After filtering, some value label categories may no longer appear in the data. Use `drop_labels()` to clean them up:

```{r}
# Subset to one region only
subset_data <- survey_data %>% filter(region == 1)

# Remove value labels for regions that are no longer in the data
clean_data <- drop_labels(subset_data, region)
val_labels(clean_data, region)
```

## Complete Example

A typical workflow for preparing SPSS data for analysis:

```{r, eval = FALSE}
# 1. Import SPSS file
data <- read_spss("survey_2024.sav")

# 2. Explore what's in the data
codebook(data)
find_var(data, "satisf")
find_var(data, "trust")

# 3. Check labels
var_label(data, q1, q2, q3)
val_labels(data, q1)

# 4. Rename variables for clarity
data <- data %>%
  rename(life_satisfaction = q1, income = q2, education = q3)

# Update variable labels
data <- var_label(data,
  life_satisfaction = "Overall life satisfaction (1-5)",
  income = "Monthly net income in euros",
  education = "Highest education level"
)

# 5. Convert categorical variables for analysis
data <- to_label(data, education, gender)

# 6. Analyze
data %>%
  describe(life_satisfaction, income, weights = sampling_weight)

data %>%
  t_test(life_satisfaction, group = gender, weights = sampling_weight)
```

## Practical Tips

1. **Keep labels as long as possible.** Labels carry important context. Only convert to factor or character when a specific function requires it.

2. **Use `find_var()` instead of `names()`.** It searches both names and labels, which is essential for SPSS datasets with non-descriptive variable names.

3. **Use `copy_labels()` after complex pipelines.** If your pipeline involves joins, reshaping, or other operations that might strip attributes, restore labels from the original data.

4. **Prefer `to_label()` over `as.factor()`.** `to_label()` uses the value labels as factor levels, giving you meaningful names instead of numeric codes.

5. **Use `set_na()` early in your workflow.** Declaring missing values immediately after import ensures they are handled correctly in all downstream analyses.

## Summary

1. `var_label()` and `val_labels()` get and set variable and value labels
2. `find_var()` searches variables by name or label pattern
3. `to_label()`, `to_character()`, `to_numeric()`, `to_labelled()` convert between formats
4. `set_na()` declares values as missing; `unlabel()` strips all metadata
5. `copy_labels()` restores labels after dplyr operations; `drop_labels()` cleans up unused labels

## Next Steps

- Transform and recode variables --- see `vignette("data-transformation")`
- Import data from SPSS, Stata, and SAS --- see `vignette("data-io")`
- Start analyzing your data --- see `vignette("descriptive-statistics")`
