---
title: "Regression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Regression analysis predicts an outcome from one or more predictors. mariposa provides two regression functions with SPSS-compatible output:

| Function | Use when... |
|----------|-------------|
| `linear_regression()` | Your outcome is continuous (e.g., income, satisfaction score) |
| `logistic_regression()` | Your outcome is binary (e.g., yes/no, high/low) |

Both functions support two interface styles:

- **Formula interface:** `linear_regression(data, y ~ x1 + x2)` -- standard R style
- **SPSS-style:** `linear_regression(data, dependent = y, predictors = c(x1, x2))` -- familiar for SPSS users

## Linear Regression

### Bivariate Regression

Start with a single predictor:

```{r}
linear_regression(survey_data, life_satisfaction ~ age)
```

The output includes four sections matching SPSS REGRESSION:

- **Model Summary**: R, R-squared, Adjusted R-squared
- **ANOVA Table**: Overall model significance
- **Coefficients**: B (unstandardized), Beta (standardized), t, p, and confidence intervals
- **Descriptives**: Mean and SD for all variables

### Understanding the Coefficients

- **B (unstandardized):** For each 1-unit increase in the predictor, the outcome changes by B units. The intercept is the predicted value when all predictors are zero.
- **Beta (standardized):** Allows comparison across predictors measured on different scales. A Beta of 0.30 means a 1-SD increase in the predictor is associated with a 0.30-SD change in the outcome.
- **p-value:** Values below 0.05 indicate that the predictor's effect is statistically significant.

### Multiple Regression

Add more predictors to explain more variance:

```{r}
linear_regression(survey_data, life_satisfaction ~ age + income + trust_government)
```

Compare Beta values to see which predictor has the strongest effect.

### SPSS-Style Interface

If you prefer the SPSS approach with separate dependent and predictor arguments:

```{r}
linear_regression(survey_data,
                  dependent = life_satisfaction,
                  predictors = c(age, income, trust_government))
```

Results are identical regardless of which interface you use.

### With Survey Weights

```{r}
linear_regression(survey_data,
                  life_satisfaction ~ age + income,
                  weights = sampling_weight)
```

Weights are treated as frequency weights, matching SPSS WEIGHT BY behavior.

### Grouped Analysis

Run separate regressions for each subgroup:

```{r}
survey_data %>%
  group_by(region) %>%
  linear_regression(life_satisfaction ~ age + income)
```

This matches SPSS SPLIT FILE BY -- a separate model is fitted for each group.

### Interpreting R-squared

R-squared tells you how much variance your predictors explain:

- R-squared = 0.01 -- 0.05: Small effect
- R-squared = 0.06 -- 0.13: Medium effect
- R-squared = 0.14+: Large effect

These benchmarks follow Cohen's (1988) conventions for the social sciences. Always check the ANOVA table to confirm the model is statistically significant.

## Logistic Regression

### When to Use

Use `logistic_regression()` when your outcome is binary. First, create a binary variable:

```{r}
survey_data <- survey_data %>%
  mutate(high_satisfaction = ifelse(life_satisfaction >= 4, 1, 0))
```

### Basic Logistic Regression

```{r}
logistic_regression(survey_data, high_satisfaction ~ age + income)
```

The output includes five sections matching SPSS LOGISTIC REGRESSION:

- **Omnibus Test**: Is the overall model significant?
- **Model Summary**: -2 Log Likelihood and pseudo R-squared values
- **Hosmer-Lemeshow Test**: Does the model fit the data well?
- **Classification Table**: How well does the model classify cases?
- **Coefficients**: B, Wald, Exp(B) (odds ratios), and confidence intervals

### Understanding Odds Ratios

The key statistic in logistic regression is Exp(B), the odds ratio:

- **Exp(B) > 1:** Each unit increase in the predictor increases the odds of the outcome
- **Exp(B) < 1:** Each unit increase in the predictor decreases the odds
- **Exp(B) = 1:** No effect

For example, Exp(B) = 1.50 means the odds increase by 50% for each 1-unit increase in the predictor. Exp(B) = 0.80 means the odds decrease by 20%.

### With Multiple Predictors

```{r}
logistic_regression(survey_data,
                    high_satisfaction ~ age + income + trust_government + education)
```

### SPSS-Style Interface

```{r}
logistic_regression(survey_data,
                    dependent = high_satisfaction,
                    predictors = c(age, income, trust_government))
```

### With Survey Weights

```{r}
logistic_regression(survey_data,
                    high_satisfaction ~ age + income,
                    weights = sampling_weight)
```

### Grouped Analysis

```{r}
survey_data %>%
  group_by(region) %>%
  logistic_regression(high_satisfaction ~ age + income)
```

### Interpreting Model Fit

**Pseudo R-squared** values are not directly comparable to linear regression R-squared, but provide a rough indication of model quality:

- **Nagelkerke R-squared**: Adjusted to reach 1.0, most commonly reported
- **Cox & Snell R-squared**: Cannot reach 1.0, always lower
- **McFadden R-squared**: Values above 0.20 indicate good fit

**Hosmer-Lemeshow Test**: A non-significant result ($p > .05$) means the model fits well. A significant result suggests the model may not adequately describe the data.

**Classification Table**: Shows the percentage of cases correctly predicted. Compare this to the base rate (percentage in the larger group) -- your model should do better than simply guessing the most common outcome.

## Complete Example

A full regression analysis workflow:

```{r}
# 1. Explore relationships first
survey_data %>%
  pearson_cor(life_satisfaction, age, income, trust_government)

# 2. Run the regression
lm_result <- linear_regression(survey_data,
                               life_satisfaction ~ age + income + trust_government,
                               weights = sampling_weight)
lm_result              # compact: R-squared + significant predictors
summary(lm_result)     # detailed: coefficients, ANOVA table, collinearity

# 3. Create binary outcome for logistic regression
survey_data <- survey_data %>%
  mutate(high_satisfaction = ifelse(life_satisfaction >= 4, 1, 0))

# 4. Run logistic regression
log_result <- logistic_regression(survey_data,
                                  high_satisfaction ~ age + income + trust_government,
                                  weights = sampling_weight)
log_result             # compact: model significance + pseudo R-squared
summary(log_result)    # detailed: odds ratios, classification table
```

## Practical Tips

1. **Check correlations first.** Use `pearson_cor()` to explore bivariate relationships before building a regression model.

2. **Examine descriptives.** Use `describe()` to check variable distributions, ranges, and missing data before running regression.

3. **Compare Beta values.** In multiple regression, standardized Beta coefficients tell you which predictor has the strongest effect, regardless of measurement scale.

4. **Report completely.** Always include R-squared, F-test (for linear) or Omnibus test (for logistic), individual coefficients, and sample size.

5. **Watch for multicollinearity.** If predictors are highly correlated with each other, coefficients may be unstable. Check bivariate correlations before interpreting results.

6. **Binary outcomes need logistic regression.** Never use `linear_regression()` with a 0/1 outcome -- predicted values can fall outside 0-1, and statistical tests are invalid.

## Summary

1. `linear_regression()` predicts continuous outcomes, providing B, Beta, ANOVA table, and R-squared
2. `logistic_regression()` predicts binary outcomes, providing odds ratios, classification table, and pseudo R-squared
3. Both support formula and SPSS-style interfaces, survey weights, and grouped analysis
4. Always check correlations and descriptives before building regression models

## Next Steps

- Create reliable scale scores first -- see `vignette("scale-analysis")`
- Compare group means directly -- see `vignette("hypothesis-testing")`
- Explore bivariate relationships -- see `vignette("correlation-analysis")`
- Handle survey weights properly -- see `vignette("survey-weights")`
