---
title: "Scale Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scale Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Scale analysis combines multiple survey items into a single score that reliably measures a concept. The typical workflow is:

1. **Check reliability** with `reliability()` --- do the items measure the same thing?
2. **Explore structure** with `efa()` --- how do items group into dimensions?
3. **Create scores** with `row_means()` --- compute mean indices
4. **Standardize** with `pomps()` --- transform to a comparable 0--100 scale

This guide uses the trust items from `survey_data` (trust in government, media, and science).

| Function | Purpose |
|----------|---------|
| `reliability()` | Internal consistency (Cronbach's Alpha) |
| `efa()` | Discover underlying dimensions (factor analysis) |
| `row_means()` | Row-wise mean indices |
| `row_sums()` | Row-wise sums |
| `row_count()` | Count specific values per row |
| `pomps()` | Percent of Maximum Possible Scores (0--100) |

## Reliability Analysis

### Basic Usage

```{r}
reliability(survey_data, trust_government, trust_media, trust_science)
```

### Detailed Output

```{r}
rel <- reliability(survey_data, trust_government, trust_media, trust_science)
summary(rel)
```

The detailed output shows item-total correlations, alpha-if-item-deleted, and inter-item correlations --- matching SPSS RELIABILITY.

### Interpreting Cronbach's Alpha

- Alpha > 0.90: Excellent
- 0.80 -- 0.90: Good
- 0.70 -- 0.80: Acceptable
- 0.60 -- 0.70: Questionable
- Below 0.60: Reconsider your items

**Item-Total Correlation** shows how well each item fits the scale. Values above 0.40 indicate good fit; below 0.20 suggests the item does not belong.

**Alpha if Item Deleted** shows what happens without each item. If alpha increases when you remove an item, that item weakens the scale.

### With Survey Weights

```{r}
reliability(survey_data, trust_government, trust_media, trust_science,
            weights = sampling_weight)
```

### Using tidyselect

```{r}
reliability(survey_data, starts_with("trust"))
```

### Grouped Analysis

Check whether reliability holds across subgroups:

```{r}
survey_data %>%
  group_by(region) %>%
  reliability(trust_government, trust_media, trust_science)
```

A scale that works well overall might be unreliable in specific subgroups. Always check when your sample spans diverse populations.

## Exploratory Factor Analysis

### Basic Usage

When you have many items, `efa()` reveals how they group into underlying dimensions:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science)
```

### Detailed Output

```{r}
efa_result <- efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science)

summary(efa_result)
```

### Understanding the Output

**KMO (Kaiser-Meyer-Olkin)** measures sampling adequacy for factor analysis:

- Above 0.80: Good to excellent
- 0.60 -- 0.80: Acceptable
- Below 0.60: Factor analysis may not be appropriate

**Bartlett's Test** should be significant ($p < .05$), confirming that meaningful correlations exist.

**Eigenvalues** show variance explained per component. By default, components with eigenvalue > 1 are retained (Kaiser criterion).

**Factor Loadings** show item-component associations:

- Above 0.70: Strong
- 0.40 -- 0.70: Moderate
- Below 0.40: Suppressed by default

### Rotation Methods

**Varimax** (default) --- assumes uncorrelated factors:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "varimax")
```

**Promax** --- allows correlated factors, produces Pattern and Structure matrices:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "promax")
```

**Oblimin** --- another oblique rotation, common in psychology:

```{r, eval = FALSE}
# Requires GPArotation package
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "oblimin")
```

### Extraction Methods

By default, `efa()` uses PCA (Principal Component Analysis). For a true factor analysis model, use Maximum Likelihood:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    extraction = "ml")
```

ML extraction provides a goodness-of-fit test and uses SMC (squared multiple correlations) as initial communalities. Combine any extraction with any rotation:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    extraction = "ml", rotation = "promax")
```

### Fixing the Number of Factors

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    n_factors = 2)
```

### With Survey Weights

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    weights = sampling_weight)
```

## Creating Scale Scores

After confirming reliability, create scores using the row operation functions. For details on `row_means()`, `row_sums()`, `row_count()`, and `pomps()`, see `vignette("data-transformation")`.

### Quick Scale Construction

```{r}
# Create mean index
survey_data <- survey_data %>%
  mutate(m_trust = row_means(., trust_government, trust_media, trust_science,
                             min_valid = 2))

# Transform to 0-100 scale
survey_data <- survey_data %>%
  mutate(trust_pomps = pomps(m_trust, scale_min = 1, scale_max = 5))

# Check the result
survey_data %>%
  describe(m_trust, trust_pomps)
```

### Using the Scale in Analysis

```{r}
# Group comparison
survey_data %>%
  t_test(m_trust, group = gender, weights = sampling_weight)
```

```{r}
# As a predictor in regression
survey_data %>%
  linear_regression(life_satisfaction ~ m_trust + age + income,
                    weights = sampling_weight)
```

## Complete Example

```{r}
# 1. Check reliability
rel <- reliability(survey_data, trust_government, trust_media, trust_science)
rel
summary(rel)

# 2. Explore factor structure
efa_result <- efa(survey_data, trust_government, trust_media, trust_science)
efa_result

# 3. Create mean index (Alpha was acceptable)
survey_data <- survey_data %>%
  mutate(m_trust = row_means(., trust_government, trust_media, trust_science,
                             min_valid = 2))

# 4. Transform to POMPS
survey_data <- survey_data %>%
  mutate(trust_pomps = pomps(m_trust, scale_min = 1, scale_max = 5))

# 5. Use in further analysis
survey_data %>%
  group_by(education) %>%
  describe(m_trust, trust_pomps, weights = sampling_weight)
```

## Practical Tips

1. **Always check reliability first.** A mean index from unreliable items produces meaningless results. Aim for Alpha > .70.

2. **Use `min_valid` wisely.** For 3--5 item scales, `min_valid = 2` is a reasonable compromise. For longer scales, require at least half the items.

3. **Specify theoretical `scale_min` / `scale_max` in `pomps()`.** Using observed values makes scores sample-dependent and non-comparable.

4. **Run EFA when you have 6+ items.** Factor analysis can reveal unexpected dimensions before you create indices.

5. **Check grouped reliability.** A scale that works well overall may be unreliable in specific subgroups (e.g., different regions or education levels).

## Summary

1. `reliability()` checks whether items form a consistent scale (Cronbach's Alpha)
2. `efa()` discovers underlying dimensions (PCA or ML extraction, Varimax/Oblimin/Promax rotation)
3. `row_means()` creates mean indices; `row_sums()` and `row_count()` provide alternatives
4. `pomps()` transforms scores to a comparable 0--100 scale
5. Always validate reliability **before** creating scale scores

## Next Steps

- Learn about row operations and transformations --- see `vignette("data-transformation")`
- Use scales in regression models --- see `vignette("regression-analysis")`
- Compare scale scores across groups --- see `vignette("hypothesis-testing")`
- Apply survey weights --- see `vignette("survey-weights")`
