---
title: "Scale Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Scale Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(mariposa)
library(dplyr)
data(survey_data)
```

## Overview

Scale analysis is the process of combining multiple survey items into a single score that reliably measures a concept. In mariposa, the typical workflow is:

1. Check reliability with `reliability()` -- do the items measure the same thing?
2. Explore structure with `efa()` -- how do items group into dimensions?
3. Create scores with `scale_index()` -- compute mean indices
4. Standardize with `pomps()` -- transform to comparable 0-100 scale

This guide walks through each step using the trust items from `survey_data` (trust in government, media, and science).

| Function | Purpose |
|----------|---------|
| `reliability()` | Check internal consistency (Cronbach's Alpha) |
| `efa()` | Discover underlying dimensions (factor analysis) |
| `scale_index()` | Create mean indices across items |
| `pomps()` | Transform to Percent of Maximum Possible Scores |

## Reliability Analysis

### Basic Usage

Before creating a scale score, check whether your items measure the same concept consistently:

```{r}
reliability(survey_data, trust_government, trust_media, trust_science)
```

The output shows Cronbach's Alpha, item-total correlations, and alpha-if-item-deleted -- matching SPSS RELIABILITY output.

### Understanding the Output

**Cronbach's Alpha** tells you how internally consistent your scale is:

- Alpha > 0.90: Excellent
- 0.80 -- 0.90: Good
- 0.70 -- 0.80: Acceptable
- 0.60 -- 0.70: Questionable
- Below 0.60: Reconsider your items

**Item-Total Correlation** shows how well each item fits:

- Above 0.40: Good fit
- 0.20 -- 0.40: Needs review
- Below 0.20: Consider removing

**Alpha if Item Deleted** shows what happens without each item. If alpha increases when you remove an item, that item weakens the scale.

### With Survey Weights

```{r}
reliability(survey_data, trust_government, trust_media, trust_science,
            weights = sampling_weight)
```

### Using tidyselect

When your items share a naming pattern, select them efficiently:

```{r}
reliability(survey_data, starts_with("trust"))
```

### Grouped Analysis

Check whether reliability holds across subgroups:

```{r}
survey_data %>%
  group_by(region) %>%
  reliability(trust_government, trust_media, trust_science)
```

## Exploratory Factor Analysis

### Basic Usage

When you have many items, `efa()` reveals how they group into underlying dimensions:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science)
```

The output includes KMO measure, Bartlett's test, eigenvalues, and the rotated component matrix -- matching SPSS FACTOR output.

### Understanding the Output

**KMO (Kaiser-Meyer-Olkin)** measures whether your data is suitable for factor analysis:

- Above 0.80: Good to excellent
- 0.60 -- 0.80: Acceptable
- Below 0.60: Factor analysis may not be appropriate

**Bartlett's Test** should be significant ($p < .05$), confirming that correlations exist in your data.

**Eigenvalues** show how much variance each component explains. By default, components with eigenvalue > 1 are retained (Kaiser criterion).

**Factor Loadings** show item-component associations:

- Above 0.70: Strong
- 0.40 -- 0.70: Moderate
- Below 0.40: Suppressed by default

### Choosing a Rotation

**Varimax** (default) assumes uncorrelated factors -- simpler to interpret:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "varimax")
```

**Oblimin** allows correlated factors -- often more realistic for social science:

```{r, eval = FALSE}
# Requires GPArotation package
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "oblimin")
```

With oblimin, you get a Pattern Matrix (unique contributions) and Structure Matrix (total correlations).

**Promax** is another oblique rotation, common in SPSS. Like oblimin, it produces Pattern and Structure matrices:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    rotation = "promax")
```

### Choosing an Extraction Method

By default, `efa()` uses PCA (Principal Component Analysis). For a true factor analysis model, use Maximum Likelihood:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    extraction = "ml")
```

ML extraction provides a **goodness-of-fit test** that evaluates whether your factor model adequately reproduces the observed correlations. A non-significant chi-square ($p > .05$) suggests good fit. ML also reports initial communalities as SMC (squared multiple correlations) instead of 1.0.

You can combine any extraction with any rotation:

```{r}
# ML extraction with Promax rotation
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    extraction = "ml", rotation = "promax")
```

### Fixing the Number of Factors

If you have a theoretical reason for a specific number of components:

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    n_factors = 2)
```

### With Survey Weights

```{r}
efa(survey_data,
    political_orientation, environmental_concern, life_satisfaction,
    trust_government, trust_media, trust_science,
    weights = sampling_weight)
```

## Creating Scale Scores

### scale_index(): Mean Indices

After confirming reliability, create a mean index:

```{r}
survey_data <- survey_data %>%
  mutate(m_trust = scale_index(., trust_government, trust_media, trust_science))

# Check the result
survey_data %>%
  describe(m_trust)
```

`scale_index()` works inside `mutate()`, computing the row-wise mean across items.

### Using tidyselect

```{r}
survey_data <- survey_data %>%
  mutate(m_trust2 = scale_index(., starts_with("trust")))
```

### Using pick()

The `pick()` function works with both `%>%` and `|>` pipes:

```{r}
survey_data <- survey_data %>%
  mutate(m_trust3 = scale_index(
    pick(trust_government, trust_media, trust_science)
  ))
```

### Minimum Valid Items

Require a minimum number of valid responses per respondent, matching SPSS `MEAN.2()` syntax:

```{r}
survey_data <- survey_data %>%
  mutate(m_trust_strict = scale_index(
    ., trust_government, trust_media, trust_science,
    min_valid = 2
  ))
```

If a respondent answered fewer than 2 items, they receive `NA` instead of a potentially unreliable score.

### pomps(): Standardize to 0-100

Transform scores to a common 0-100 scale for cross-scale comparison:

```{r}
survey_data <- survey_data %>%
  mutate(trust_pomps = pomps(m_trust, scale_min = 1, scale_max = 5))

survey_data %>%
  describe(trust_pomps)
```

A score of 0 means the minimum possible, 100 means the maximum. Always specify the theoretical scale range for consistent results.

### Multiple Variables at Once

```{r}
survey_data <- survey_data %>%
  mutate(across(
    c(trust_government, trust_media, trust_science),
    ~ pomps(.x, scale_min = 1, scale_max = 5),
    .names = "{.col}_pomps"
  ))
```

## Complete Example

A typical scale analysis workflow from start to finish:

```{r}
# 1. Check reliability
rel <- reliability(survey_data, trust_government, trust_media, trust_science)
rel                # compact: Alpha + interpretation
summary(rel)       # detailed: item statistics, inter-item correlations

# 2. Explore factor structure
efa_result <- efa(survey_data, trust_government, trust_media, trust_science)
efa_result         # compact: KMO + variance explained
summary(efa_result, communalities = FALSE)  # detailed, skip communalities

# 3. Create mean index (Alpha was acceptable)
survey_data <- survey_data %>%
  mutate(m_trust = scale_index(., trust_government, trust_media, trust_science))

# 4. Transform to POMPS for reporting
survey_data <- survey_data %>%
  mutate(trust_pomps = pomps(m_trust, scale_min = 1, scale_max = 5))

# 5. Use in further analyses
survey_data %>%
  t_test(m_trust, group = region, weights = sampling_weight)
```

## Practical Tips

1. **Always check reliability first.** A mean index from unreliable items produces meaningless results. Aim for Alpha > .70.

2. **Use `min_valid` wisely.** For scales with 3-5 items, `min_valid = 2` is a reasonable compromise between data loss and score quality.

3. **Specify `scale_min` and `scale_max` in `pomps()`.** Using observed values instead of theoretical range makes scores incomparable across samples.

4. **Run EFA when you have many items.** With 6+ items, factor analysis can reveal unexpected dimensions before you create indices.

5. **Check grouped reliability.** Alpha may differ substantially across subgroups (e.g., regions, education levels). A scale that works well overall might be unreliable in specific groups.

## Summary

1. `reliability()` checks whether items form a consistent scale (Cronbach's Alpha)
2. `efa()` discovers underlying dimensions in your items (Principal Component Analysis)
3. `scale_index()` computes mean indices for use in further analyses
4. `pomps()` transforms scores to a comparable 0-100 scale

## Next Steps

- Use your new scale scores in group comparisons -- see `vignette("hypothesis-testing")`
- Explore relationships between scales -- see `vignette("correlation-analysis")`
- Predict outcomes with regression -- see `vignette("regression-analysis")`
- Apply survey weights to all analyses -- see `vignette("survey-weights")`
