---
title: "Introduction to mariposa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mariposa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(mariposa)
library(dplyr)
```

## Overview

mariposa (*Marburg Initiative for Political and Social Analysis*) makes it easy to analyze survey data in R. Whether you are working with social surveys, customer satisfaction data, or market research, this package helps you get reliable results quickly.

The package handles the complexities of survey analysis for you --- things like sampling weights, grouped comparisons, and missing data are all taken care of automatically. Every function is validated against SPSS, so you can trust the numbers.

## Getting Started

Let's start with the included example data. It contains responses from 2,500 people about their life satisfaction, demographics, and opinions.

```{r}
data(survey_data)
glimpse(survey_data)
```

## Understanding Your Data

### Summary Statistics

The `describe()` function gives you a complete summary of numeric variables. The `weights` argument ensures results represent the population correctly:

```{r}
survey_data %>%
  describe(age, income, weights = sampling_weight)
```

The output shows:

- **n**: Number of valid responses
- **mean**: The average ($\bar{x}$)
- **sd**: Standard deviation --- how spread out the values are
- **median**: The middle value (50th percentile)
- **range**: Minimum and maximum values

### Categorical Variables

For categorical data like education level or employment status, use `frequency()`:

```{r}
survey_data %>%
  frequency(education, weights = sampling_weight)
```

This tells you how many people fall into each category, what percentage they represent, and cumulative percentages.

## Comparing Groups

### Two Groups: t-Test

Want to know if men and women differ in life satisfaction? Use `t_test()`:

```{r}
survey_data %>%
  t_test(life_satisfaction, group = gender, weights = sampling_weight)
```

If the p-value is below 0.05, the difference is statistically significant. But always check the effect size too --- a significant result with a tiny effect might not be practically meaningful.

### Three or More Groups: ANOVA

When comparing more than two groups, use `oneway_anova()`:

```{r}
survey_data %>%
  oneway_anova(life_satisfaction, group = education, weights = sampling_weight)
```

## Finding Relationships

### Correlation

Are age and income related? Pearson's *r* measures linear association:

```{r}
survey_data %>%
  pearson_cor(age, income, weights = sampling_weight)
```

- **r > 0**: As one variable increases, the other tends to increase
- **r < 0**: As one increases, the other tends to decrease
- **r near 0**: No linear relationship

### Cross-Tabulation

See how two categorical variables relate:

```{r}
survey_data %>%
  crosstab(education, employment, weights = sampling_weight)
```

## Working with Groups

### Regional Analysis

Use `group_by()` from dplyr to compare statistics across subgroups:

```{r}
survey_data %>%
  group_by(region) %>%
  describe(income, life_satisfaction, weights = sampling_weight)
```

### Multiple Variables at Once

Test several variables simultaneously:

```{r}
survey_data %>%
  t_test(trust_government, trust_media, trust_science,
         group = gender, weights = sampling_weight)
```

## Why Weights Matter

Survey weights make your results representative of the population. Here is the difference they make:

```{r}
unweighted <- mean(survey_data$age, na.rm = TRUE)

weighted_result <- w_mean(survey_data, age, weights = sampling_weight)
weighted <- weighted_result$results$weighted_mean

cat("Sample average age:", round(unweighted, 1), "\n")
cat("Population average age:", round(weighted, 1), "\n")
cat("Difference:", round(weighted - unweighted, 1), "years\n")
```

Always use weights when they are provided --- unweighted results can be misleading.

## A Typical Workflow

Here is how a complete analysis might look:

```{r}
# 1. Descriptive overview
survey_data %>%
  describe(life_satisfaction, weights = sampling_weight)

# 2. Compare across groups
survey_data %>%
  group_by(education) %>%
  describe(life_satisfaction, weights = sampling_weight)

# 3. Test for significant differences
result <- survey_data %>%
  oneway_anova(life_satisfaction, group = education,
               weights = sampling_weight)
print(result)

# 4. If significant, find out which groups differ
tukey_test(result)
```

## Tips for Success

1. **Always start with descriptives.** Understand your data before running tests.
2. **Check your assumptions.** Skewness near 0 suggests roughly normal data; values beyond $\pm 1$ indicate notable skew.
3. **Use appropriate tests.** Continuous data: t-test, ANOVA, Pearson. Ordinal data: Mann-Whitney, Spearman. Categorical data: chi-square.
4. **Report effect sizes.** Statistical significance alone does not tell you whether a difference matters in practice.

## Quick Reference

| Task | Function | Example |
|------|----------|---------|
| Summarize numeric | `describe()` | `describe(age, income)` |
| Count categories | `frequency()` | `frequency(education)` |
| Compare 2 groups | `t_test()` | `t_test(income, group = gender)` |
| Compare 3+ groups | `oneway_anova()` | `oneway_anova(income, group = education)` |
| Test association | `chi_square()` | `chi_square(education, employment)` |
| Measure correlation | `pearson_cor()` | `pearson_cor(age, income)` |
| Cross-tabulate | `crosstab()` | `crosstab(education, region)` |

All functions support survey weights via `weights =` and grouped analysis via `group_by()`.

## Next Steps

- `vignette("descriptive-statistics")` --- Deep dive into summaries and frequencies
- `vignette("hypothesis-testing")` --- Comparing groups and testing differences
- `vignette("correlation-analysis")` --- Measuring relationships
- `vignette("survey-weights")` --- Working with weighted data
